---
title: "Using gisaimsr data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using gisaimsr data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo = FALSE}
# links
oagb <- "https://www.github.com/open-AIMS/"
oaio <- "https://open-AIMS.github.io/"
rgh <- "https://www.github.com/ropensci/"
rio <- "https://ropensci.github.io/"
gagb <- paste0(oagb, "gisaimsr/")
gais <- paste0(gagb, "issues/new")
gaio <- paste0(oaio, "gisaimsr/")
gard <- paste0(gaio, "index.html")
garf <- paste0(gaio, "reference/index.html")

dagb <- paste0(rgh, "dataaimsr/")
daio <- paste0(rio, "dataaimsr/")

cranpkgs <- "https://cran.r-project.org/package="
sf <- paste0(cranpkgs, "sf")
rt <- paste0(cranpkgs, "remotes")
gs <- paste0(cranpkgs, "ggspatial")
gp <- paste0(cranpkgs, "ggplot2")
```

## Introduction

`gisaimsr` was developed to provide scientists with easy access to some key
spatial datasets associated with key marine ecosystems in Australia where most
of the work conducted by the Australian Institute of Marine Science takes 
place. If you would like to see more spatial datasets made available through 
this package, please submit a request via our [issues][gais] page on GitHub.

Different from most R packages, this package should be viewed as a data 
package only. Therefore, the purpose of this vignette is to teach how one can: 
(*i*) query which spatial datasets are available and how to access them, and
(*ii*) provide illustrated examples of data exploration and plots.

Our datasets were originally provided by Geoscience Australia (GA) and the 
Great Barrier Reef Marine Park Authority (GBRMPA). In September 2021, we
transformed all data to the GDA2020 coordinate system. Please see our 
[README][gard] page for full information regarding data and package licenses.

[gard]: `r gard`
[gais]: `r gais`

## Installing the package

Before moving forward on this page, please refer to our [README][gard] page 
for full information regarding package installation and dependencies.

## The datasets

Our datasets are in [simple features data.frame][sf] format, which is being
actively developed and soon is likely to be the main go-to spatial object
in R. There is no point in repeating abundant information; if you want an
overview of spatial data objects in R, and simple features in particular,
please see the [further readings](#further-readings) section at the end of 
this vignette.

Currently all of our spatial datasets are from Queensland because of AIMS 
historical focus on the Great Barrier Reef. If you would like to access 
spatial datasets from other parts of the country (e.g. tropical Western 
Australia and Northern Territory), please submit a request via our
[issues][gais] page on GitHub. A list of datasets and their associated full 
documentation can be found on our [reference](garf) page, but briefly we 
provide datasets on:

* Great Barrier Reef Marine Park Boundary (object `gbr_feat`)

* Great Barrier Reef Features (object `gbr_bounds`)

* Great Barrier Reef Marine Park Zoning (object `gbr_zoning`)

* Natural Resource Management Marine Regions Relating to the Great Barrier Reef Marine Park (object `nrm_regions`)

* Marine Water Bodies within the Great Barrier Reef (object `wbodies`)

* Queensland Hydrology data (object `qld_hydro`)

All datasets are standardised to the Geocentric Datum of Australia
([GDA94][gda]). However, this vignette will guide you through some examples 
about changing the projection of the datasets if needed.

[sf]: `r sf`
[garf]: `r garf`
[gda]: http://www.geoproject.com.au/gda.faq.html

## Examples

Before we explore and plot some data, let's first load some packages that we 
are going to need further down the track; most of them can be downloaded using 
the `install.packages` function in R. [`gisaimsr`][gaio] and
[`dataaimsr`][daio] can be installed using the [`remotes`][rt] package
(see more on our [README][gard] page).

```{r, message = FALSE}
library(raster)
library(tidyverse)
library(ggspatial)
library(sf)
library(dataaimsr)
library(gisaimsr)
library(ggrepel)
```

We recommend always loading the `sf` package whenever loading `gisaimsr`---this
will facilitate the summary view of the datasets. We look at what the 
`gbr_feat` object comprises:

```{r}
gbr_feat
```

A two-part summary was printed to the screen. The top part summarises the 
features (rows) X fields (columns) of the spatial dataset, its geographical 
boundaries and projection. The second part displays the first 10 rows of the 
dataset. These fields contain all of the original info on the dataset, and the 
last column `geometry` contains the actual spatial information of that 
particular feature, e.g.

```{r}
gbr_feat[1, "geometry"]
```

The nice thing about this type of format is that it can be subset in whichever 
way we want, as we would do for a regular data.frame. For instance, we can 
isolate the first occurrence of the dataset

```{r}
gbr_feat[1, ]
```

In a similar fashion, we can subset the data based on a particular field value

```{r}
gbr_feat %>%
  filter(GBR_NAME == "Forrester Reef")
```

which is extremely handy when we want to plot particular sections of the 
dataset

```{r}
gbr_feat %>%
  data.frame %>%
  select(FEAT_NAME) %>%
  distinct()
```

```{r, fig.width = 5, fig.height = 5}
gbr_feat %>%
  dplyr::filter(FEAT_NAME != "Mainland") %>%
  ggplot(data = .) +
    geom_sf() +
    theme_classic()
```

As we can see the default of `geom_sf` is not necessarily nice. We can play
with some of the arguments in the function to make the plots more pleasant, 
including colouring the land based on a particular field (i.e. column), and
adding some ocean colour to the plot's background

```{r, fig.width = 5, fig.height = 5}
ggplot(data = gbr_feat) +
  geom_sf(mapping = aes(fill = FEAT_NAME), lwd = 0.01) +
  theme_classic() +
  theme(panel.background = element_rect(fill = "lightblue",
                                        colour = "lightblue",
                                        size = 0.5, linetype = "solid")) +
  labs(x = "Longitude",
       y = "Latitude",
       title = "The Great Barrier Reef",
       subtitle = "by geographic feature") +
  scale_fill_brewer(name = "Type", palette = "Spectral")
```

We can also overlay different objects from `gisaimsr`, clip the map to a 
particular area, and add points to our map

```{r, fig.width = 7, fig.height = 5}
coords <- data.frame(reef = c("Heron reef", "Cockatoo reef"), lat = c(-23.44, -20.70), lon = c(151.92, 150.99))
ggplot() +
  geom_sf(data = gbr_zoning, mapping = aes(fill = TYPE), colour = NA) +
  geom_sf(data = gbr_feat, lwd = 0.01, fill = "grey90") +
  annotation_scale(location = "bl", width_hint = 0.5) +
  annotation_north_arrow(location = "bl", which_north = "true",
      pad_x = unit(0.25, "in"), pad_y = unit(0.4, "in"),
      style = north_arrow_fancy_orienteering) +
  coord_sf(xlim = c(148, 154.5), ylim = c(-25, -18)) +
  geom_point(data = coords, mapping = aes(x = lon, y = lat)) +
  geom_label_repel(data = coords, 
                   mapping = aes(x = lon, y = lat,
                                 label = reef),
                   hjust = 0, vjust = 0) +
  theme_classic() +
  labs(x = "Longitude",
       y = "Latitude",
       title = "Southern Great Barrier Reef",
       subtitle = "by zone") +
  scale_fill_brewer(name = "Zone", palette = "Spectral")
```

The functions `annotation_scale` and `annotation_north_arrow` from the 
[`ggspatial`][gs] package allow the addition of scale bars and orientation.
`coord_sf` from [`ggplot2`][gp] allows for the delimitation of the plot area.
It also allows one to change the projection of the map by changing the 
Coordinate Reference System (CRS)

```{r, fig.width = 5, fig.height = 5}
ggplot(data = gbr_feat) +
  geom_sf(lwd = 0.01, fill = "burlywood2") +
  theme_classic() +
  coord_sf(crs = "+proj=moll") +
  theme(panel.background = element_rect(fill = "lightblue",
                                        colour = "lightblue",
                                        size = 0.5, linetype = "solid")) +
  labs(x = "Longitude",
       y = "Latitude",
       title = "The Great Barrier Reef",
       subtitle = "World Mollweide projection")
```

The argument `crs` in `coord_sf` could have been an EPSG (European Petroleum 
Survey Group) code (e.g. `crs = sf::st_crs(7844)` for the new
[GDA2020][gda2020]). One can search for different projections and codes
[here](https://spatialreference.org/ref/epsg/?page=1) and
[here](https://epsg.io/).

One can also crop a particular dataset for a user-defined region
before plotting the object. This requires some code to first create a bounding
box on equivalent CRS to the target dataset to be cropped, and then some
wrangling to fix any original invalid geometries from the original dataset:

```{r, fig.width = 5, fig.height = 5, warning = FALSE, message = FALSE}
# first define box
cropbox <- extent(140, 146, -15, -10) %>%
  as("SpatialPolygons") %>%
  st_as_sf %>%
  st_set_crs(st_crs(gbr_feat))

gbr_feat %>%
  st_make_valid %>%
  st_intersection(cropbox) %>%
  ggplot(data = .) +
    geom_sf(lwd = 0.01, fill = "burlywood2") +
    theme_classic() +
    theme(panel.background = element_rect(fill = "lightblue",
                                          colour = "lightblue",
                                          size = 0.5, linetype = "solid")) +
    labs(x = "Longitude",
         y = "Latitude",
         title = "Northern Great Barrier Reef")
```

To finalise, we also provide visualisation of the remaining datasets available
in `gisaimsr`:

```{r, fig.width = 5, fig.height = 5}
ggplot() +
  geom_sf(data = gbr_bounds, fill = "darkred", colour = NA) +
  geom_sf(data = gbr_feat, lwd = 0.01, fill = "burlywood2") +
  annotation_scale(location = "bl", width_hint = 0.5) +
  annotation_north_arrow(location = "bl", which_north = "true",
      pad_x = unit(0.25, "in"), pad_y = unit(0.4, "in"),
      style = north_arrow_fancy_orienteering) +
  theme_classic() +
  theme(panel.background = element_rect(fill = "lightblue",
                                        colour = "lightblue",
                                        size = 0.5, linetype = "solid")) +
  labs(x = "Longitude",
       y = "Latitude",
       title = "The Great Barrier Reef",
       subtitle = "Marine park boundary")
```

```{r, fig.width = 6, fig.height = 5}
ggplot() +
  geom_sf(data = nrm_regions,
          mapping = aes(fill = NAME), lwd = 0.01) +
  theme_classic() +
  labs(x = "Longitude",
       y = "Latitude",
       title = "The Great Barrier Reef",
       subtitle = "by NRM marine regions") +
  scale_fill_brewer(name = "Region", palette = "Spectral")
```

```{r, fig.width = 6, fig.height = 5}
ggplot() +
  geom_sf(data = wbodies,
          mapping = aes(fill = MarineWate), lwd = 0.01) +
  theme_classic() +
  labs(x = "Longitude",
       y = "Latitude",
       title = "The Great Barrier Reef",
       subtitle = "by marine water bodies") +
  scale_fill_brewer(name = "Type of body", palette = "Spectral")
```

```{r, fig.width = 5, fig.height = 5}
ggplot() +
  geom_sf(data = qld_hydro, lwd = 0.01, colour = "dodgerblue3") +
  theme_classic() +
  labs(x = "Longitude",
       y = "Latitude",
       title = "Queensland hydrology map")
```

[gaio]: `r gaio`
[daio]: `r daio`
[rt]: `r rt`
[gs]: `r gs`
[gp]: `r gp`
[gda2020]: http://www.ga.gov.au/scientific-topics/positioning-navigation/geodesy/datums-projections/gda2020

## Further readings

* Spatial data in R:

    - https://www.jessesadler.com/post/gis-with-r-intro/

    - https://www.flutterbys.com.au/stats/tut/tut5.4.html

* Simple features data.frame:

    - https://r-spatial.github.io/sf/articles/sf1.html

    - https://www.jessesadler.com/post/simple-feature-objects/

    - https://github.com/rstudio/cheatsheets/raw/master/sf.pdf

* Plotting spatial data in R:

    - https://www.r-spatial.org/r/2018/10/25/ggplot2-sf.html

    - https://cran.r-project.org/web/packages/rnaturalearth/vignettes/rnaturalearth.html

* About Coordinate Reference Systems:

    - https://www.nceas.ucsb.edu/sites/default/files/2020-04/OverviewCoordinateReferenceSystems.pdf
